---
title: "Referencing other files"
author: "Jenna Krall"
date: "October 29, 2014"
output: html_document
---


# Required packages

Required packages from CRAN: 

1. knitr


# Read in the external code
```{r rc}
library(knitr)
read_chunk("ex2_filereference.R")
```




# Can call code using chunk name

Now we have a bunch of text, say for a manuscript or report.  And we don't want hundreds of lines of R code dividing up our text-- it makes it difficult to read and difficult to edit.  Instead, we can call external .R files where our code lives, but embed them here.


```{r gendat, echo = T, tidy = T}
```

See?  it is much easier to read.  Oh and as an aside, we can also include LaTeX equation editing.  So, our linear model is $y_i = \beta_0 + \beta_1 x_i + \epsilon_i$ where $\epsilon\sim \text{N}(0, 1)$



# In-line code

```{r}
x <- rnorm(100)
```

We can also show in-line code.  After taking 100 samples from a standard normal distribution, we computed a mean of `r round(mean(x), 2)` and a standard deviation of `r sd(x)`.  

The mean of x is 0.137.

